---
title: "Big Data Project"
author: "Ryne"
date: "2023-11-03"
output: html_document
---

#Loading Libraries
```{r}
library(arrow);library(sparklyr);library(dplyr);library(ggplot2);library(dbplot);library(corrr);library(lubridate);library(plotly);library(doParallel);library(plumber)
```


#Connection with Spark
```{r}
sc <- spark_connect(master = "local", version = "3.4.0")
sleep_csv <- spark_read_csv(sc,
                        name = "Sleep",
                        path = "Sleep_Efficiency.csv")

parquet_path <- "Sleep_parquet"
spark_write_parquet(sleep_csv, path = parquet_path, mode = "overwrite")
sleep <- spark_read_parquet(sc, path = parquet_path, overwrite = TRUE)
```


#EDA

##cleaning step
```{r}
#looking at our data
head(sleep)

#using glimpse we find out some data that we want to clean
glimpse(sleep)

#this step cleans by
#1. dropping na rows
#2. dropping the ID row
#3. changing smoking status and gender as binary variables
#4. setting bedtime and wakeup_time as hour in int format
#named as sleep_clean
sleep_clean <- sleep |>
               filter_all(all_vars(!is.na(.)))|>
               select(-ID) |>
               mutate(Smoking_status = as.integer(ifelse(Smoking_status == "Yes", 1, 0)),
                      Awakenings = as.integer(Awakenings),
                      Gender = as.integer(ifelse(Gender == "Female", 1, 0)),
                      Bedtime = (hour(Bedtime)),
                      Wakeup_time = (hour(Wakeup_time)))

glimpse(sleep_clean)
head(sleep_clean)
```


##Data Exploration

###Summary Statistics
```{r}
#checking for abnomalies summary statistics
sleep_clean|>
  sdf_describe()
```


###Frequency distribution of variables
```{r}
# Frequency distribution of numerical variables
sleep_clean_df <- sleep_clean |>
                  collect()

for (col in names(sleep_clean_df)){
  graphs <- ggplot(sleep_clean_df, aes_string(x = col)) +
    geom_histogram(binwidth = 0.15, fill = "skyblue", color = "black")+
    labs(title = paste("Histogram of", col), x = col, y = "Frequency")

  print(graphs)
}
```


###Correlations of variables
```{r}
# #getting correlation of all variables
# sleep_corr <- sleep_clean |>
#                 correlate(use = "pairwise.complete.obs", method = "pearson") |>
#                 shave(upper = TRUE)
# 
# sleep_corr
# 
# corr_plot <- rplot(sleep_corr, print_cor = TRUE) +
#                 scale_x_discrete(guide = guide_axis(angle = 45))
# 
# corr_plot
```

###relationship between target and some interesting predictors

```{r}
#investigating relationship between target and individual predictors

#for smoke against sleep efficiency
smoke_prop_data <- sleep_clean |>
                   group_by(Smoking_status) |>
                   summarise(
                      avg_sleep_efficiency = mean(Sleep_efficiency, na.rm = TRUE),
                      sd_sleep_efficiency = sd(Sleep_efficiency, na.rm = TRUE),
                      n = n()) |>
                   mutate(se = sd_sleep_efficiency/sqrt(n)) |>
                    collect()
smoke_prop_data
```

```{r}
#prepping for plot
average_sleep_eff_df <- sleep_clean |>
                        summarise(avg_sleep_efficiency = mean(Sleep_efficiency))|>
                        collect()

average_sleep_eff_vec <- average_sleep_eff_df$avg_sleep_efficiency

#plotting relationship between smoking status and sleep efficiency
smoke_prop_data |>
  mutate(Smoking_status = as.factor(Smoking_status))|>
  ggplot(aes(x = Smoking_status, y = avg_sleep_efficiency))+
  geom_point(size = 2)+
  geom_errorbar(aes(ymin = avg_sleep_efficiency - 1.96*se,
                    ymax = avg_sleep_efficiency + 1.96*se),
                    width = 0.1) +
  geom_hline(yintercept = average_sleep_eff_vec, linetype = "dashed")+
  labs(title = "Average sleep efficiency based on smoking status")+
  ylab("Average sleep efficiency")+
  xlab("Smoking status")
```

```{r}
#investigating relationship between alcohol consumption and sleep efficiency
alcohol_prop_data <- sleep_clean %>%
  group_by(Alcohol_consumption) %>%
  summarise(
    avg_sleep_efficiency = mean(Sleep_efficiency, na.rm = TRUE),
    sd_sleep_efficiency = sd(Sleep_efficiency, na.rm = TRUE),
    n = n()
  ) %>%
  mutate(se = sd_sleep_efficiency/sqrt(n)) %>%
  arrange(Alcohol_consumption) %>%
  collect()

alcohol_prop_data

#plot of alcohol consumption against sleep efficiency
alc_con <- sleep_clean %>%
  ggplot(aes(x = Alcohol_consumption, y = Sleep_efficiency)) +
  geom_point(aes(Alcohol_consumption)) +
  labs(title = "Sleep Efficiency vs. Alcohol Consumption",
       x = "Alcohol Consumption (oz)",
       y = "Sleep Efficiency (%)") +
  theme_minimal()

alc_con
```

```{r}
#investigating relationship between caffeine consumption and sleep efficiency
caffeine_prop_data <- sleep_clean %>%
  group_by(Caffeine_consumption) %>%
  summarise(
    avg_sleep_efficiency = mean(Sleep_efficiency, na.rm = TRUE),
    sd_sleep_efficiency = sd(Sleep_efficiency, na.rm = TRUE),
    n = n()
  ) %>%
  mutate(se = sd_sleep_efficiency/sqrt(n)) %>%
  arrange(Caffeine_consumption) %>%
  collect()

caffeine_prop_data

#plot of caffeine consumption against sleep efficiency
caff_con <- sleep_clean %>%
  ggplot(aes(x = Caffeine_consumption, y = Sleep_efficiency)) +
  geom_point(aes(Caffeine_consumption)) +
  labs(title = "Sleep Efficiency vs. Caffeine Consumption",
       x = "Caffeine Consumption (times)",
       y = "Sleep Efficiency (%)") +
  theme_minimal()

caff_con
```

```{r}
#relationship of gender against sleep efficiency
gender_prop_data <- sleep_clean |>
                    group_by(Gender) |>
                    summarise(avg_sleep_efficiency = mean(Sleep_efficiency),
                              sd_sleep_efficiency = sd(Sleep_efficiency),
                              n = n())|>
                    mutate(se = sd_sleep_efficiency/sqrt(n)) |>
                    collect()
gender_prop_data
```

```{r}
#plot of gender against sleep efficiency
gender_prop_data |>
  mutate(Gender = as.factor(Gender)) |>
  ggplot(aes(x = Gender, y = avg_sleep_efficiency))+
  geom_point(size = 2)+
  geom_errorbar(aes(ymin = avg_sleep_efficiency - 1.96*se,
                    ymax = avg_sleep_efficiency + 1.96*se),
                    width = 0.1) +
  geom_hline(yintercept = average_sleep_eff_vec, linetype = "dashed")+
  labs(title = "Average sleep efficiency based on gender")+
  ylab("Average sleep efficiency")+
  xlab("Gender")
```

```{r}
#for exercise frequency against sleep efficiency
exercise_prop_data <- sleep_clean |>
                    group_by(Exercise_frequency) |>
                    summarise(avg_sleep_efficiency = mean(Sleep_efficiency),
                              sd_sleep_efficiency = sd(Sleep_efficiency),
                              n = n())|>
                    mutate(se = sd_sleep_efficiency/sqrt(n)) |>
                    arrange(Exercise_frequency) |>
                    collect()
exercise_prop_data
```

##Visualisation
```{r}
#visualising light sleep percentage against sleep_efficiency using raster
sleep_clean |>
  dbplot_raster(x = Light_sleep_percentage, y = Sleep_efficiency, fill = n(), resolution = 10)

#raster plot for deep sleep percentage against sleep efficiency using raster
sleep_clean |>
  dbplot_raster(x = Deep_sleep_percentage, y = Sleep_efficiency, fill = n(), resolution = 10)

#raster plot for REM sleep against sleep efficiency using rastor plot
sleep_clean |>
  dbplot_raster(x = REM_sleep_percentage, y = Sleep_efficiency, fill = n(), resolution = 10)

#raster plot for Sleep Duration against sleep efficiency using rastor plot
sleep_clean |>
  dbplot_raster(x = Sleep_duration, y = Sleep_efficiency, fill = n(), resolution = 10)
```

```{r}
grouped_sleep8 <- sleep_clean |>
                  group_by(Bedtime)|>
                  summarise(n = n(), ave_sleep = mean(Sleep_efficiency))|>
                  arrange(desc(ave_sleep))|>
                  collect()
bedtime_plot <- grouped_sleep8 |>
                mutate(Bedtime = as.factor(Bedtime)) |>
                ggplot(aes(x=Bedtime, y=ave_sleep))+
                geom_point(aes(size = n))+
                labs(title = "The effects of Bedtime on Sleep efficiency")+
                ylab("Average sleep efficiency")+
                xlab("Bedtime in 24 Hours")+
                theme_minimal()
ggplotly(bedtime_plot)
```

```{r}
# visualising the frequency of sleep duration using histogram
sleep_clean |>
  dbplot_histogram(Sleep_duration, binwidth = 1) +
  labs(title = "Sleep Duration Distribution",
       subtitle = "Histogram of Sleep Duration") +
  xlab("Sleep Duration")
```

```{r}
sleep_clean |>
  dbplot_raster(x = Sleep_duration, y = Sleep_efficiency, fill = n(), resolution = 10)
```

```{r}
# boxplot of Deep sleep percentage with respect to Smoking status
sleep_clean |>
  ggplot(aes(x = as.factor(Smoking_status), y = Deep_sleep_percentage)) +
  geom_boxplot() +
  xlab("Smoking Status") +
  ylab("Deep Sleep Percentage")

# boxplot of Deep sleep percentage with respect to Gender
sleep_clean |>
  ggplot(aes(x = as.factor(Gender), y = Deep_sleep_percentage)) +
  geom_boxplot() +
  xlab("Gender") +
  ylab("Deep Sleep Percentage")
```

```{r}
#average sleep efficiency by age
grouped_sleep <- sleep_clean |>
                group_by(Age) |>
                summarise(avg_sleep_efficiency = mean(Sleep_efficiency))|>
                collect()
grouped_sleep |>
  ggplot(aes(x = Age, y = avg_sleep_efficiency)) +
  geom_line()+
  labs(title = "Average sleep efficiency by age")+
  ylab("Average sleep efficiency")+
  xlab("Age")+
  coord_cartesian(ylim = c(0.4, 1))
```

```{r}
# boxplot of Sleep efficiency with respect to Alcohol consumption
sleep_clean |>
  ggplot(aes(x = as.factor(Alcohol_consumption), y = Sleep_efficiency)) +
  geom_boxplot() +
  xlab("Alcohol consumption") +
  ylab("Sleep efficiency")

# boxplot of Sleep efficiency with respect to REM sleep percentage
# rational: Alcohol consumption harms REM sleep, which is important for memory and learning
sleep_clean |>
  ggplot(aes(x = as.factor(REM_sleep_percentage), y = Sleep_efficiency)) +
  geom_boxplot() +
  xlab("REM sleep percentage") +
  ylab("Sleep efficiency")

# boxplot of Sleep efficiency with respect to Caffeine consumption
sleep_clean |>
  ggplot(aes(x = as.factor(Caffeine_consumption), y = Sleep_efficiency)) +
  geom_boxplot() +
  xlab("Caffeine consumption") +
  ylab("Sleep efficiency")
```

```{r}
# boxplot of Deep sleep percentage against Awakenings
# rational: To see if the number of awakenings has an effect on deep sleep
# the higher the frequency of awakenings, sleep quality decreases
sleep_clean |>
  ggplot(aes(x = as.factor(Awakenings), y = Deep_sleep_percentage)) +
  geom_boxplot() +
  xlab("Awakenings") +
  ylab("Deep sleep percentage")
```

```{r}
# Alcohol consumption and smoking status
grouped_sleep10 <- sleep_clean |>
                   mutate(Alcohol_status = ifelse(Alcohol_consumption %in% c(1,2,3,4,5),1,0))|>
                   group_by(Alcohol_status, Smoking_status)|>
                   summarise(n = n())|>
                   group_by(Alcohol_status)|>
                   summarize(count = sum(n), prop = sum(Smoking_status * n) / sum(n))|> #prop of those who smoke
                   collect()

grouped_sleep10 |>
  mutate(Alcohol_status = as.factor(Alcohol_status))|>
  ggplot(aes(x = Alcohol_status, y = prop))+
  geom_col(width = 0.3)+
  labs(title = "Proportion of smokers who consume alcohol")+
  xlab("Alcohol consumption status")+
  ylab("Proportion")
```


#Modelling

##Split data into training and testing
```{r}
# Split data into training and testing set
sleep_split <- sleep_clean |>
               sdf_random_split(training = 0.8,
                                testing = 0.2,
                                seed = 1337)

sleep_train <- sleep_split$training
sleep_test <- sleep_split$testing

# There are 325 observations in sleep_train
sleep_train |>
  summarise(count())
attach(sleep_train)
sleep_clean |>
  summarise(count())
```

##Model 1
```{r}
# ml1 <- sleep_train |>
#        ml_linear_regression(formula = Sleep_efficiency ~ . -Bedtime -Wakeup_time)|>
#         tidy()
# ml1
```


##Model 2
```{r}
#removing all insignificant variables
ml2 <- sleep_train |>
       ml_linear_regression(formula = Sleep_efficiency ~ . -Bedtime -Wakeup_time-Gender-Sleep_duration-Deep_sleep_percentage-Caffeine_consumption)|>
       tidy()
ml2
```

##Model 3
```{r}
#adding caffiene consumption
ml3 <- sleep_train |>
       ml_linear_regression(formula = Sleep_efficiency ~ . -Bedtime -Wakeup_time-Gender-Sleep_duration-Deep_sleep_percentage)|>
       tidy()

ml3
```

##Model 1 Cross Validation
```{r}
pipeline1 <- ml_pipeline(sc) |>
             ft_r_formula(formula = Sleep_efficiency ~ . -Bedtime -Wakeup_time)|>
             ml_linear_regression()

grid <- list(linear_regression = list(elastic_net_param = 0, reg_param = 0))

cv <- ml_cross_validator(
  sc,
  estimator = pipeline1,
  evaluator = ml_regression_evaluator(
              sc,
              label_col = "Sleep_efficiency"),
  estimator_param_maps = grid,
  num_folds = 10,
  parallelism = 1,
  seed = 1337
)

pipeline_model1 <- ml_fit(cv,sleep_train)
rmse1 <- ml_validation_metrics(pipeline_model1);rmse1
```


##Model 2 Cross Validation
```{r}
pipeline2 <- ml_pipeline(sc) |>
             ft_r_formula(formula = Sleep_efficiency ~ . -Bedtime -Wakeup_time-Gender-Sleep_duration-Deep_sleep_percentage-Caffeine_consumption)|>
             ml_linear_regression()

grid <- list(linear_regression = list(elastic_net_param = 0, reg_param = 0))

cv <- ml_cross_validator(
  sc,
  estimator = pipeline2,
  evaluator = ml_regression_evaluator(
              sc,
              label_col = "Sleep_efficiency"),
  estimator_param_maps = grid,
  num_folds = 10,
  parallelism = 1,
  seed = 1337
)

pipeline_model2 <- ml_fit(cv,sleep_train)
rmse2 <- ml_validation_metrics(pipeline_model2);rmse2
```

##Model 3 Cross Validation
```{r}
pipeline3 <- ml_pipeline(sc) |>
             ft_r_formula(formula = Sleep_efficiency ~ . -Bedtime -Wakeup_time-Gender-Sleep_duration-Deep_sleep_percentage)|>
             ml_linear_regression()

grid <- list(linear_regression = list(elastic_net_param = 0, reg_param = 0))

cv <- ml_cross_validator(
  sc,
  estimator = pipeline3,
  evaluator = ml_regression_evaluator(
              sc,
              label_col = "Sleep_efficiency"),
  estimator_param_maps = grid,
  num_folds = 10,
  parallelism = 1,
  seed = 1337
)

pipeline_model3 <- ml_fit(cv,sleep_train)
rmse3 <- ml_validation_metrics(pipeline_model3);rmse3
```

##RMSE plot
```{r}
rmse_df <- data.frame(model = c(1,2,3),
                      rmse = c(rmse1$rmse,rmse2$rmse,rmse3$rmse)) |>
           mutate(model = as.factor(model))

rmse_plot <- rmse_df |>
            ggplot(aes(x = model, y = rmse, group = 1)) +
            geom_point() +
            geom_line(color = "deepskyblue") +
            geom_text(aes(label = round(rmse,5)), vjust = -0.5)+
            labs(title = "Plot of RMSE by Model", x = "Model", y = "RMSE")

rmse_plot
```

##Creating vectors for pipeline
```{r}
all_str_variables <- sleep |>
                     select_if(is.character)|>
                     names()

all_variables_ml3 <- ml3$term[-1]
str_variables_ml3 <- intersect(all_str_variables, all_variables_ml3)
num_variables_ml3 <- setdiff(all_variables_ml3, str_variables_ml3)
```

##Creating pipeline with for loop
```{r}
pipeline <- ml_pipeline(sc) |>
            ft_vector_assembler(input_col = num_variables_ml3,
                                output_col = "features") |>
            ft_standard_scaler(input_col = "features",
                               output_col = "stdz_features",
                               with_mean = TRUE)

encoder_input_vec <- c()
for (variable in str_variables_ml3) {
  output_col <- paste0(variable, "_indexed")
  pipeline <- pipeline |>
                        ft_string_indexer(
                        input_col = variable,
                        output_col = paste0(variable, "_indexed"))
  encoder_input_vec <- c(encoder_input_vec, output_col)
}


encoder_output_vec <- c()
for (variable in encoder_input_vec) {
  output_col <- paste0(sub("_indexed", "", variable), "_encoded")
  pipeline <- pipeline |>
                        ft_one_hot_encoder(
                        input_cols = variable,
                        output_col = paste0(sub("_indexed", "", variable), "_encoded"))
  encoder_output_vec <- c(encoder_output_vec , output_col)
}

input_vector <- c("stdz_features", encoder_output_vec)

pipeline <- pipeline |>
            ft_vector_assembler(
            input_cols =  input_vector,
            output_col = "final_features") |>

            ml_linear_regression(
            features_col = "final_features",
            label_col = "Sleep_efficiency")

pipeline
```

##Cross Validation to tune hyperparameters
```{r}
ncores <- detectCores(logical = FALSE)
grid <- list(linear_regression = list(elastic_net_param = c(0, 1), reg_param = c(0.001, 0.01) ))

cv <- ml_cross_validator(
  sc,
  estimator = pipeline,
  estimator_param_maps = grid,
  evaluator = ml_regression_evaluator(sc,
                                      label_col = "Sleep_efficiency"),
  num_folds = 10,
  parallelism = ncores,
  seed = 1337
)
```

##Employ CV estimator to training set
```{r}
cv_model <- ml_fit(cv,sleep_train);cv_model
```

##Inspect cross validation result
```{r}
ml_validation_metrics(cv_model) |>
  arrange(rmse)
```

##Best model and its coefficient
```{r}
ml_stage(cv_model$best_model, stage = "linear_regression")
ml_stage(cv_model$best_model, stage = "linear_regression")$coefficients
```

##Using the best model to predict sleep efficienct on test set
```{r}
predictions <- ml_transform(cv_model, sleep_test)|>
               collect()
glimpse(predictions)
```

##Predicted vs Actual sleep efficiency plot
```{r}
predictions |>
  ggplot(aes(x = prediction, y = Sleep_efficiency)) +
  geom_point(color = "deepskyblue4", size = 2) +
  geom_abline(color = "deepskyblue2", linetype = "dashed") +
  labs(
    x = "Predicted sleep efficiency",
    y = "Actual sleep efficiency",
    title = "Predicted vs Actual sleep efficiency"
  )
```

#Plumblr
```{r}
# # api.R
# #' Predict Sleep Efficiency
# #' @param Bedtime Numeric
# #' @param Wakeup_time Numeric
# #' @param Gender Numeric
# #' @param Sleep_duration Numeric
# #' @param Deep_sleep_percentage Numeric
# #' @get /predict
# function(Bedtime, Wakeup_time, Gender, Sleep_duration, Deep_sleep_percentage) {
#   new_data <- data.frame(Bedtime = as.numeric(Bedtime), Wakeup_time = as.numeric(Wakeup_time), Gender = as.numeric(Gender), Sleep_duration = as.numeric(Sleep_duration), Deep_sleep_percentage = as.numeric(Deep_sleep_percentage))
#   prediction <- predict(pipeline_model3, new_data)
#   list(sleep_efficiency = prediction)
# }
```
```{r}
saveRDS(cv_model, "cv_model.rds")
```

```{r}
r <- plumb("plumber.R")
r$run(port = 8000)
```
```{r}
class(cv_model)
```


