---
title: "Group 1's Big Data Project on Sleep Efficiency"
output: html_document
---

## Loading library
```{r}
library(arrow);library(sparklyr);library(dplyr);library(ggplot2);library(dbplot);library(corrr);library(doParallel);library(plotly);library(plumber)
```

## Connection with Spark
```{r}
sc <- spark_connect(master = "local", version = "3.4.0")
sleep_csv <- spark_read_csv(sc,
                        name = "Sleep",
                        path = paste0("file://", getwd(), "/Sleep_Efficiency.csv"))

parquet_path <- "Sleep_parquet"
spark_write_parquet(sleep_csv, path = parquet_path, mode="overwrite")
sleep <- spark_read_parquet(sc, path = parquet_path)
```

## Exploratory Data Analysis
## Data Cleaning
```{r}
#looking at our data
head(sleep)

#using glimpse we find out some data that we want to clean
glimpse(sleep)

#this step cleans by
#1. dropping na rows
#2. dropping the ID row
#3. changing smoking status and gender as binary variables
#4. setting bedtime and wakeup_time as hour in int format
#named as sleep_clean
sleep_clean <- sleep |>
               filter_all(all_vars(!is.na(.)))|>
               select(-ID) |>
               mutate(Smoking_status = as.integer(ifelse(Smoking_status == "Yes", 1, 0)),
                      Awakenings = as.integer(Awakenings),
                      Gender = as.integer(ifelse(Gender == "Female", 1, 0)),
                      Bedtime = (hour(Bedtime)),
                      Wakeup_time = (hour(Wakeup_time)))

glimpse(sleep_clean)
head(sleep_clean)
```

## Data Exploration
### Summary statistics
```{r}
#checking for anomalies in summary statistics
sleep_clean|>
  sdf_describe()
```

### Frequency distribution of numerical variables
```{r}
sleep_clean_df <- sleep_clean |>
                  collect() 
  
for (col in names(sleep_clean_df)){
  graphs <- ggplot(sleep_clean_df, aes_string(x = col)) + 
    geom_histogram(binwidth = 0.15, fill = "skyblue", color = "black")+
    labs(title = paste("Histogram of", col), x = col, y = "Frequency")
  
  print(graphs)
}
```

### Correlation of variables
```{r}
sleep_corr <- sleep_clean |>
                correlate(use = "pairwise.complete.obs", method = "pearson") |>
                shave(upper = TRUE)

sleep_corr

corr_plot <- rplot(sleep_corr, print_cor = TRUE) +
             scale_x_discrete(guide = guide_axis(angle = 45))

corr_plot
```

```{r}
grouped_sleep1 <- sleep_clean |>
                   mutate(Alcohol_status = ifelse(Alcohol_consumption %in% c(1,2,3,4,5),1,0))|>
                   group_by(Alcohol_status, Smoking_status)|>
                   summarise(n = n())|>
                   group_by(Alcohol_status)|>
                   summarize(count = sum(n), prop = sum(Smoking_status * n) / sum(n))|> #prop of those who smoke
                   collect()

grouped_sleep1 |> 
  mutate(Alcohol_status = as.factor(Alcohol_status))|>
  ggplot(aes(x = Alcohol_status, y = prop))+
  geom_col(width = 0.3)+
  labs(title = "Proportion of smokers who consume alcohol")+
  xlab("Alcohol consumption status")+
  ylab("Proportion")
```

### Chi-square test for independence for alcohol consumption and smoking status
```{r}
contingency_table <- sleep_clean|>
                     mutate(Alcohol_status = as.integer(ifelse(Alcohol_consumption %in% c(1,2,3,4,5),1,0)))|>
                     sdf_crosstab("Smoking_status", "Alcohol_status")|>
                     collect()|>
                     sapply(as.numeric)|>
                     as.matrix()

contingency_matrix <- matrix(c(60,111,60,94), nrow = 2)
chisq.test(contingency_matrix)
```
Since p value = 0.5436 > 0.05, do not reject null hypothesis, smoking status and alcohol status are independent

### Relationship between target and some interesting predictors
#### Smoking_status vs sleep efficiency
```{r}
smoke_prop_data <- sleep_clean |>
                   group_by(Smoking_status) |>
                   summarise(
                      avg_sleep_efficiency = mean(Sleep_efficiency, na.rm = TRUE),
                      sd_sleep_efficiency = sd(Sleep_efficiency, na.rm = TRUE),
                      n = n()) |>
                   mutate(se = sd_sleep_efficiency/sqrt(n)) |>
                    collect()
smoke_prop_data
```

```{r}
#prepping for plot
average_sleep_eff_df <- sleep_clean |>
                        summarise(avg_sleep_efficiency = mean(Sleep_efficiency))|>
                        collect()

average_sleep_eff_vec <- average_sleep_eff_df$avg_sleep_efficiency

#plotting relationship between smoking status and sleep efficiency
smoke_prop_data |>
  mutate(Smoking_status = as.factor(Smoking_status))|>
  ggplot(aes(x = Smoking_status, y = avg_sleep_efficiency))+
  geom_point(size = 2)+
  geom_errorbar(aes(ymin = avg_sleep_efficiency - 1.96*se,
                    ymax = avg_sleep_efficiency + 1.96*se),
                    width = 0.1) +
  geom_hline(yintercept = average_sleep_eff_vec, linetype = "dashed")+
  labs(title = "Average sleep efficiency based on smoking status")+
  ylab("Average sleep efficiency")+
  xlab("Smoking status")
```

#### Gender vs Sleep efficiency
```{r}
gender_prop_data <- sleep_clean |>
                    group_by(Gender) |>
                    summarise(avg_sleep_efficiency = mean(Sleep_efficiency),
                              sd_sleep_efficiency = sd(Sleep_efficiency),
                              n = n())|>
                    mutate(se = sd_sleep_efficiency/sqrt(n)) |>
                    collect()
gender_prop_data

gender_prop_data |>
  mutate(Gender = as.factor(Gender)) |>
  ggplot(aes(x = Gender, y = avg_sleep_efficiency))+
  geom_point(size = 2)+
  geom_errorbar(aes(ymin = avg_sleep_efficiency - 1.96*se,
                    ymax = avg_sleep_efficiency + 1.96*se),
                    width = 0.1) +
  geom_hline(yintercept = average_sleep_eff_vec, linetype = "dashed")+
  labs(title = "Average sleep efficiency based on gender")+
  ylab("Average sleep efficiency")+
  xlab("Gender")
```
#### Age vs Sleep efficiency
```{r}
grouped_sleep2 <- sleep_clean |> 
                  group_by(Age) |>
                  summarise(avg_sleep_efficiency = mean(Sleep_efficiency))|>
                  collect() 
grouped_sleep2 |>
  ggplot(aes(x = Age, y = avg_sleep_efficiency)) +
  geom_line()+
  labs(title = "Average sleep efficiency by age")+
  ylab("Average sleep efficiency")+
  xlab("Age")+
  coord_cartesian(ylim = c(0.4, 1))
```

#### Sleep duration vs age
```{r}
grouped_sleep3 <- sleep_clean |> 
                group_by(Age) |>
                summarise(avg_sleep_duration = mean(Sleep_duration))|>
                collect() 
grouped_sleep3 |>
  ggplot(aes(x = Age, y = avg_sleep_duration)) +
  geom_line()+
  labs(title = "Average sleep duration by age")+
  ylab("Average sleep duration")+
  xlab("Age")
```
#### Awakenings vs Sleep efficiency
```{r}
sleep_clean |>
  ggplot(aes(x = as.factor(Awakenings), y = Sleep_efficiency)) +
  geom_boxplot() +
  xlab("Awakenings") +
  ylab("Sleep efficiency")
```

#### Caffeine consumption vs sleep efficiency
```{r}
sleep_clean |>
  dbplot_raster(x = Caffeine_consumption, y = Sleep_efficiency, fill = n(), resolution = 10)
```

#### Caffeine consumption vs awakenings
```{r}
grouped_sleep4 <- sleep_clean |>
                  group_by(Awakenings)|>
                  summarise(Caffeine_consumption_percentile = percentile(Caffeine_consumption, array(0.25,0.5,0.75)))|>
                  mutate(Caffeine_consumption_percentile = explode(Caffeine_consumption_percentile))|>
                  collect()

grouped_sleep4

grouped_sleep4 |>
  mutate(Awakenings = as.factor(Awakenings))|>
  ggplot(aes(x = Awakenings, y = Caffeine_consumption_percentile))+
  geom_boxplot()+
  labs(title = "The effects of Caffeine consumption on Awakenings")+
  ylab("Caffeine consumption")+
  xlab("Awakenings")
```

#### Alcohol consumption vs sleep efficiency
```{r}
alcohol_prop_data <- sleep_clean |>
  group_by(Alcohol_consumption) |>
  summarise(
    avg_sleep_efficiency = mean(Sleep_efficiency, na.rm = TRUE),
    sd_sleep_efficiency = sd(Sleep_efficiency, na.rm = TRUE),
    n = n()
  ) |>
  mutate(se = sd_sleep_efficiency/sqrt(n)) |>
  arrange(Alcohol_consumption) |>
  collect()

alcohol_prop_data


alcohol_prop_data |>
  ggplot(aes(x = Alcohol_consumption, y = avg_sleep_efficiency))+
  geom_point(size = 2)+
  geom_errorbar(aes(ymin = avg_sleep_efficiency - 1.96*se,
                    ymax = avg_sleep_efficiency + 1.96*se),
                    width = 0.1) +
  geom_hline(yintercept = average_sleep_eff_vec, linetype = "dashed")+
  labs(title = "Average sleep efficiency based alcohol consumption")+
  ylab("Average sleep efficiency")+
  xlab("Alcohol consumption (in oz)")
```

#### Exercise_frequency against sleep_efficiency

```{r}
exercise_prop_data <- sleep_clean |>
                    group_by(Exercise_frequency) |>
                    summarise(avg_sleep_efficiency = mean(Sleep_efficiency),
                              sd_sleep_efficiency = sd(Sleep_efficiency),
                              n = n())|>
                    mutate(se = sd_sleep_efficiency/sqrt(n)) |>
                    arrange(Exercise_frequency) |>
                    collect()
exercise_prop_data

#plotting relationship between exercise and sleep efficiency
exercise_prop_data |>
  ggplot(aes(x = Exercise_frequency, y = avg_sleep_efficiency))+
  geom_point(size = 2)+
  geom_errorbar(aes(ymin = avg_sleep_efficiency - 1.96*se,
                    ymax = avg_sleep_efficiency + 1.96*se),
                    width = 0.1) +
  geom_hline(yintercept = average_sleep_eff_vec, linetype = "dashed")+
  labs(title = "Average sleep efficiency based on exercise")+
  ylab("Average sleep efficiency")+
  xlab("Exercise Frequency (per week)")
```

#### Relationship between Bedtime and sleep efficiency
```{r}
grouped_sleep5 <- sleep_clean |>
                  group_by(Bedtime)|>
                  summarise(n = n(), ave_sleep = mean(Sleep_efficiency))|>
                  arrange(desc(ave_sleep))|>
                  collect()

bedtime_plot <- grouped_sleep5 |>
                mutate(Bedtime = as.factor(Bedtime)) |>
                ggplot(aes(x=Bedtime, y=ave_sleep))+
                geom_point(aes(size = n))+
                labs(title = "The effects of Bedtime on Sleep efficiency")+
                ylab("Average sleep efficiency")+
                xlab("Bedtime in 24 Hours")+
                theme_minimal()
ggplotly(bedtime_plot)
```

#### Relationship between Bedtime, Sleep duration and sleep efficiency
```{r}
grouped_sleep6 <- sleep_clean |>
                  group_by(Bedtime, Sleep_duration)|>
                  summarise(n = n(), ave_sleep = mean(Sleep_efficiency))|>
                  arrange(desc(ave_sleep))|>
                  filter(ave_sleep>0.79 & Sleep_duration>6) |>
                  collect()

bedtime_plot <- grouped_sleep6 |>
                mutate(Bedtime = as.factor(Bedtime)) |>
                ggplot(aes(x=Bedtime, y=ave_sleep))+
                geom_point(aes(size = n))+
                geom_point(aes(colour= Sleep_duration))+
                labs(title = "The effects of Bedtime on Sleep efficiency")+
                ylab("Average sleep efficiency")+
                xlab("Bedtime in 24 Hours")+
                theme_minimal()
ggplotly(bedtime_plot)
```


# Modelling
### Spliting data into training and test set
```{r}
# Split data into training and testing set
sleep_split <- sleep_clean |>
               sdf_random_split(training = 0.8,
                                testing = 0.2,
                                seed = 1337)

sleep_train <- sleep_split$training
sleep_test <- sleep_split$testing
```
### Model 1
```{r}
ml1 <- sleep_train |> 
       ml_linear_regression(formula = Sleep_efficiency ~ . -Bedtime -Wakeup_time)|>
        tidy()
ml1
# Gender, sleep_duration, deep_sleep, caffeine_consumption are not significant
```

### Model 2 - Removing all insignificant variables
```{r}
ml2 <- sleep_train |> 
       ml_linear_regression(formula = Sleep_efficiency ~ . -Bedtime -Wakeup_time-Gender-Sleep_duration-Deep_sleep_percentage-Caffeine_consumption)|>
       tidy()
ml2
```

### Model 3 - Adding caffeine consumption
```{r}
ml3 <- sleep_train |> 
       ml_linear_regression(formula = Sleep_efficiency ~ . -Bedtime -Wakeup_time-Gender-Sleep_duration-Deep_sleep_percentage)|>
       tidy()

ml3
```


# Model 1 CV
```{r}
pipeline1 <- ml_pipeline(sc) |>
             ft_r_formula(formula = Sleep_efficiency ~ . -Bedtime -Wakeup_time)|> 
             ml_linear_regression()

grid <- list(linear_regression = list(elastic_net_param = 0, reg_param = 0))

cv <- ml_cross_validator(
  sc,
  estimator = pipeline1,
  evaluator = ml_regression_evaluator(
              sc, 
              label_col = "Sleep_efficiency"),
  estimator_param_maps = grid,
  num_folds = 10,
  parallelism = 1,
  seed = 1337
)

pipeline_model1 <- ml_fit(cv,sleep_train)
rmse1 <- ml_validation_metrics(pipeline_model1);rmse1
```
# Model 2 CV
```{r}
pipeline2 <- ml_pipeline(sc) |>
             ft_r_formula(formula = Sleep_efficiency ~ . -Bedtime -Wakeup_time-Gender-Sleep_duration-Deep_sleep_percentage-Caffeine_consumption)|> 
             ml_linear_regression()

grid <- list(linear_regression = list(elastic_net_param = 0, reg_param = 0))

cv <- ml_cross_validator(
  sc,
  estimator = pipeline2,
  evaluator = ml_regression_evaluator(
              sc, 
              label_col = "Sleep_efficiency"),
  estimator_param_maps = grid,
  num_folds = 10,
  parallelism = 1,
  seed = 1337
)

pipeline_model2 <- ml_fit(cv,sleep_train)
rmse2 <- ml_validation_metrics(pipeline_model2);rmse2
```
# Model 3 CV
```{r}
pipeline3 <- ml_pipeline(sc) |>
             ft_r_formula(formula = Sleep_efficiency ~ . -Bedtime -Wakeup_time-Gender-Sleep_duration-Deep_sleep_percentage)|> 
             ml_linear_regression()

grid <- list(linear_regression = list(elastic_net_param = 0, reg_param = 0))

cv <- ml_cross_validator(
  sc,
  estimator = pipeline3,
  evaluator = ml_regression_evaluator(
              sc, 
              label_col = "Sleep_efficiency"),
  estimator_param_maps = grid,
  num_folds = 10,
  parallelism = 1,
  seed = 1337
)

pipeline_model3 <- ml_fit(cv,sleep_train)
rmse3 <- ml_validation_metrics(pipeline_model3);rmse3
```

### RMSE plot
```{r}
rmse_df <- data.frame(model = c(1,2,3), 
                      rmse = c(rmse1$rmse,rmse2$rmse,rmse3$rmse)) |>
           mutate(model = as.factor(model))

rmse_plot <- rmse_df |>
            ggplot(aes(x = model, y = rmse, group = 1)) +
            geom_point() +  
            geom_line(color = "deepskyblue") + 
            geom_text(aes(label = round(rmse,5)), vjust = -0.5)+
            ylim(0.058, 0.06)+
            labs(title = "Plot of RMSE by Model", x = "Model", y = "RMSE")  

rmse_plot
```


### Creating vectors for pipeline
```{r}
all_str_variables <- sleep |> 
                     select_if(is.character)|>
                     names()

all_variables_ml3 <- ml3$term[-1]
str_variables_ml3 <- intersect(all_str_variables, all_variables_ml3)
num_variables_ml3 <- setdiff(all_variables_ml3, str_variables_ml3)

```

### Create pipeline with for loop
```{r}
pipeline <- ml_pipeline(sc) |>
            ft_vector_assembler(input_col = num_variables_ml3,
                                output_col = "features") |>
            ft_standard_scaler(input_col = "features",
                               output_col = "stdz_features",
                               with_mean = TRUE) 

encoder_input_vec <- c()
for (variable in str_variables_ml3) {
  output_col <- paste0(variable, "_indexed")
  pipeline <- pipeline |>
                        ft_string_indexer(
                        input_col = variable, 
                        output_col = paste0(variable, "_indexed")) 
  encoder_input_vec <- c(encoder_input_vec, output_col)
}


encoder_output_vec <- c()
for (variable in encoder_input_vec) {
  output_col <- paste0(sub("_indexed", "", variable), "_encoded")
  pipeline <- pipeline |>
                        ft_one_hot_encoder(
                        input_cols = variable, 
                        output_col = paste0(sub("_indexed", "", variable), "_encoded"))
  encoder_output_vec <- c(encoder_output_vec , output_col)
}

input_vector <- c("stdz_features", encoder_output_vec)

pipeline <- pipeline |>
            ft_vector_assembler(
            input_cols =  input_vector, 
            output_col = "final_features") |>
            
            ml_linear_regression(
            features_col = "final_features", 
            label_col = "Sleep_efficiency")

pipeline
```

### Creating pipeline; same output as the above pipeline with for-loops
```{r}
pipeline <- ml_pipeline(sc) |>
            ft_vector_assembler(input_cols = c("Age", "REM_sleep_percentage", "Light_sleep_percentage", "Awakenings", "Caffeine_consumption", "Alcohol_consumption", "Exercise_frequency"),
                                output_col = "features") |>
            ft_standard_scaler(input_col = "features",
                               output_col = "stdz_features",
                               with_mean = TRUE) |>
            ft_string_indexer(input_col = "Smoking_status",
                              output_col = "Smoking_status_indexed")|>
            ft_one_hot_encoder(input_cols = "Smoking_status_indexed",
                               output_cols = "Smoking_status_encoded")|>
            ft_vector_assembler(input_cols = c("stdz_features", "Smoking_status_encoded"),
                                output_col = "final_features")|>
            ml_linear_regression(features_col = "final_features",
                                 label_col = "Sleep_efficiency")

pipeline
```

### Cross validation to tune hyperparameters for our 3rd model
```{r}
ncores <- detectCores(logical = FALSE)
grid <- list(
        linear_regression = list(
        elastic_net_param = seq(from = 0, to = 1, length = 11), 
        reg_param = seq(from = 0.001, to = 0.01, length = 20)))

cv <- ml_cross_validator(
  sc,
  estimator = pipeline,
  estimator_param_maps = grid,
  evaluator = ml_regression_evaluator(sc, 
                                      label_col = "Sleep_efficiency"),
  num_folds = 10,
  parallelism = ncores,
  seed = 1337
)
```

### Employ CV estimator to training set
```{r}
cv_model <- ml_fit(cv,sleep_train);cv_model
```

### Inspect cross validation results
```{r}
ml_validation_metrics(cv_model) |>
              arrange(rmse)
```
### Best model and its coefficients
```{r}
ml_stage(cv_model$best_model, stage = "linear_regression")

bestmodel_coeff <- ml_stage(cv_model$best_model, stage = "linear_regression")$coefficients
bestmodel_coeff
```
### Using the best model to predict sleep efficiency on test set
```{r}
predictions <- ml_transform(cv_model, sleep_test)|>
               collect()
glimpse(predictions)
```

### Actual vs predicted sleep efficiency plot
```{r}
predictions |> 
  ggplot(aes(x = prediction, y = Sleep_efficiency)) +
  geom_point(color = "deepskyblue4", size = 2) +
  geom_abline(color = "deepskyblue2", linetype = "dashed") +
  labs(
    x = "Predicted sleep efficiency",
    y = "Actual sleep efficiency",
    title = "Predicted vs Actual sleep efficiency"
  )
```
### Saving the best model to disk
```{r}
ml_save(cv_model$best_model, path = "Sleep_efficiency_model", overwrite = TRUE)
```
### Plumber

```{r}
plumb(file = "Sleep_efficiency_plumber.R") |>
  pr_run(port = 8000)
```
